---
title: "ACGCA"
author: "Michael Fell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## ACGCA Introduction
The Allometrically Constrained Growth and Carbon Allocation (ACGCA) model is an individually based model of tree growth and mortality originally developed by Ogle and Pacala (2009). As the name implies the ACGCA model makes use of functional traits and allometries between tissue types (i.e. leaves, roots, and stems) to determine carbon allocation in a tree. 

Carbon dynamics within the tree are essential to this model. Carbon is allocated between a structural and two storage pools in a way that satisfies the aforementioned allometric relationships (Ogle and Pacala (2009). One storage pool is associated with leaf and fine root tissue. Labile carbon in this pool is retranslocated when the tissue senesceses (Ogle and Pacala 2009). The other storage pool is in the sapwood and is used in times of stress (high labile carbon demand relative to carbon inputs). Finally there is a transient labile carbon pool representative of recent photosynthate and retranslocated carbon. It is not associated with a specific tissue and is immediately redistributed within the tree to accommodate biomass production, respiration, and allocation to storage. If the labile carbon available to the tree is depleted the tree dies (Ogle and Pacala 2009).

There are six possible physiological states in the ACGCA model: healthy, static, shrinking, recovering, recovered, or dead (Ogle and Pacala 2009). Each state is described by a set of difference equations satisfying mass-balance relationshipts between structural and labile carbon pools. The only environmental driver included in the model is parmax (annual radiation for a site). 

For a full description of the model see Ogle and Pacala (2009) and the accompanying supplemental material. The model has also been used in several other studies (Fell et al. 2018 and Gemoets et al 2013).

## Using the Model
To use the model requires setting up a vector of parameters. The following parameter values are taken from Ogle and Pacala (2009), and are based on inputs to the C version of the code translated from Matlab by Darren Gemoets. Units for each parameter can be found in Ogle and Pacala (2009), Fell el al. (2018), and the runacgca function documentation. Fell et al. (2018). is published in Ecosphere, an open source journal, and thus should be available to all users of this code.

```{r, echo=FALSE}
# A code block to show the values for acru taken from Ogle and Pacala (2009)

load("../data/acru.rda")
load("../data/parms.rda")

# disable scientific notation
options(scipen = 999) # default is 1

acru.data <- cbind(parms, acru)
library(knitr)
library(kableExtra)
knitr::kable(data.frame(parameter.names=parms, parameter.values=acru), "html") %>%
  kableExtra::kable_styling(full_width=FALSE, position = "left")

# return to default setting
options(scipen = 0)
```

Once the data are loaded they can be used to call the ACGCA model. There are a few other parameters that can be set. The value of parmax (2060 MJ m^-2^ year^-1^) is taken from Ogle and Pacala (2009) and is full sunlight for a site in North Carolina. There are 801 data points in the time series for radius because the time step defaults to 16 steps per year. The extra point is for initilization. **NOTE: If the radius is set too low (any lower than it is set here) the tree will not grow**.

```{r}
# Call the ACGCA model (runacgca)
acgcaout <- ACGCA::runacgca(sparms = acru, r0 = 0.0054, parmax = 2060*0.5, years = 50, fulloutput = FALSE)

# Plot the radius and height output from the model.
par(mar=c(4,4,1,1))
plot(x=1:801/16, y = acgcaout$r, type="l", xlab = "time (years)", ylab = "radius (m)")
plot(x=1:801/16, y = acgcaout$h, type="l", xlab = "time (years)", ylab = "height (m)")
```

Other parameters can be set as outlined in the function documentation for "runacgca."

## Light Environment
Along with running the ACGCA model under a fixed annual light value it is possible to have changing light conditions over the course of the simulation. If the desired effect is to have light change each year a vector can be supplied for the "parmax" input parameter. The first value in the vector is the initilization value with the following values corresponding to the simulation post initilization. If the desire is to have values for each year and a 1/16 of a year time step is used (steps = 16), then elements 2 through 17 in the vector should be set to the value for year 1. 

```{r}
# Set up a light vector for a 50 year simulation where light decreases 10% per decade
decrease <- seq(from = 1, to = 0.2, by = -0.2)
decrease <- rep(x = decrease, each = 10*16) # repeat each value years * steps times

# create a light vector with length years * steps + 1
parmax <- c(0, rep(2060, 50*16)*decrease)

# run the ACGCA model
acgcaout <- ACGCA::runacgca(sparms = acru, r0 = 0.0054, parmax = parmax, years = 50, steps = 16, fulloutput = FALSE)

# make a plot of the radius over time and the growth state over time.
par(mar=c(4,4,1,1))
plot(x= 1:801/16, y=acgcaout$r, xlab="years", ylab="radius (m)", type="l")
plot(x= 1:801/16, y=acgcaout$growth_st, xlab="years", ylab="growth state", type="l")
```

In the second graph the growth states correspond to 1 (healthy), 4 (static), and 5 (shrinking) states. Thus, the tree was healthy for the first 30 years and then stoped growing and then transitioned to a shrinking state. It is possible if this simulation continued the tree would run out of carbon and die.

## Gap Simulations
NOTE: GAP SIMULATIONS ARE CURRENTLY EXPERIMENTAL AND THERE ARE SOME KNOWN ISSUES. THIS SHOULD NOT BE USED FOR RESEARCH PURPOSES AT THIS TIME.

Gap simulations similar to those described in Ogle and Pacala (2009) and Fell et al. (2018) can be easily implemented. To use gap simulations "gapsim" should be set to TRUE and gapvars needs to be assigned a list defining how gap dynamics will progress. The list should have three elements "list=(gt, ct, tbg)." The first variable gap time (gt) gives the length of time in years the gap is open. The second variable defines the closure time (ct) or the number of years it takes for the forest canopy to transition from open to closed. The final value is the time between gaps (tbg) which corresponds to the total time from when a gap forms to when it opens again. For example, if gt=50, ct=10, and tbg=200 (the arbitrary default), then the canopy would be open for 50 years, close over 10 years and be closed for the remaining 140 years.

```{r}
# run the ACGCA model
gapvariables <- list(gt=2, ct=3, tbg=20)
acgcaout <- ACGCA::runacgca(sparms = acru, r0 = 0.0054, parmax = 2060, years = 100, steps = 16, gapvars=gapvariables, gapsim=TRUE, fulloutput = TRUE)

# make a plot of the radius over time and the growth state over time.
par(mar=c(4,4,1,1))
plot(x= 1:1601/16, y=acgcaout$r, xlab="years", ylab="radius (m)", type="l")
plot(x= 1:1600/16, y=acgcaout$growth_st[1:1600], xlab="years", ylab="growth state", type="l")
plot(x= 1:1600/16, y=acgcaout$APARout[1:1600], xlab="years", ylab="light", type="l")
```

## References:
Fell, M., J. Barber, J. W. Lichstein, and K. Ogle. 2018. Multidimensional trait space informed by a mechanistic model of tree growth and carbon allocation. Ecosphere 9. DOI:10.1002/ecs2.2060

Gemoets, D., J. Barber, and K. Ogle. 2013. Reversible jump MCMC for inference in a deterministic individual-based model of tree growth for studying forest dynamics. Environmetrics 24:433-448.

Ogle, K., and S. W. Pacala. 2009. A modeling framework for inferring tree growth and allocation from physiological, morphological and allometric traits. Tree Physiology 29:587-605.
